---
title: "Running an N-of-1 study end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running an N-of-1 study end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(nof1kit)
```

A single-case (N-of-1) experiment has four stages: design a randomization
schedule, collect data against it, check what came back, and analyze it. Packages
for single-case analysis are plentiful. The first three stages are usually
hand-rolled, which is where the errors live.

This vignette walks all four, using a real study: 70 days, three prompts a day,
melatonin randomized nightly. The data shipped with the package are the ones
actually collected.

## 1. Design

`design_schedule()` produces a balanced sequence with a cap on how many identical
days may run consecutively. Long runs of the same condition confound the
intervention with time, so a run-length constraint is standard practice in N-of-1
designs. The seed is required, not optional: a preregistered schedule has to be
reproducible from what is written in the preregistration.

```{r}
sched <- design_schedule(n_days = 70, max_run = 2, seed = 20260218)
head(sched, 12)
```

The sampling is exact rather than by rejection. For a balanced 70-day design with
`max_run = 2`, fewer than one permutation in a million satisfies the constraint,
so drawing until one fits is not workable. Instead the number of valid completions
from each intermediate state is counted by dynamic programming, and the sequence
is drawn day by day in proportion to those counts. The result is a uniform draw
over all valid schedules: no admissible sequence is more likely than any other.
That is a stronger property than blocked randomization, which restricts the
sample space in ways that are rarely stated.

`check_schedule()` reports what a reviewer, or a preregistration reader, would
want to see.

```{r}
check_schedule(sched)
```

Balanced counts, no run longer than two, and a negative lag-1 autocorrelation,
which is what the run cap buys: the sequence alternates more than chance would.

## 2. Hand the schedule to the collection tool

`write_schedule()` writes the day-to-condition mapping as JSON.

```{r}
f <- tempfile(fileext = ".json")
write_schedule(sched, f)
cat(substr(readLines(f), 1, 90))
```

That format is what the companion iOS app and Shortcut both read, so a schedule
designed here can be loaded onto a phone without an intermediate step. Generation
and execution stay separate on purpose: an instrument that can also rewrite the
randomization is an instrument that can silently unblind a study.

## 3. Read what came back, and check it

The study data are bundled with the package. Real files rarely use the column name
you expect, so `read_ema()` takes the name of the timestamp column.

```{r, error = TRUE}
path <- system.file("extdata", "melatonin_ema.csv", package = "nof1kit")
ema <- read_ema(path, start_date = "2026-02-18", timestamp_col = "datetime")
```

That error is not a demonstration. It is a real defect in a real dataset, found by
writing this vignette.

The file carries its own `study_day` column, and for 68 of 195 records that column
disagrees with the calendar. Reading with a `start_date` makes the two indices
meet, so the conflict surfaces immediately. Reading without one accepts the file's
index, and `validate_ema()` catches it instead:

```{r}
ema <- read_ema(path, timestamp_col = "datetime")

v <- validate_ema(
  ema,
  ranges = list(mood = c(0, 100), agency = c(0, 100),
                metacognition = c(0, 100), melatonin = c(0, 1)),
  n_days = 70
)
v
subset(v$issues, check == "inconsistent_study_day")
```

One study day covering two calendar dates shifts every record after it. Here the
shift runs for the final 27 days of the study, and it is invisible to the checks
most pipelines actually perform: the values are all in range, no timestamps are
duplicated, nothing is missing, and every `study_day` sits inside `1:70`. Only
comparing the index against the calendar reveals it.

The result is data, not a printed report, so a pipeline that must not proceed on
bad input can act on it:

```{r, error = TRUE}
stopifnot(v$n_issues == 0)
```

## 4. Compliance, and why the definition matters

Compliance in EMA studies is usually reported as records divided by prompts. On
these data that gives:

```{r}
records <- nrow(ema)
prompts <- 70 * 3
round(100 * records / prompts, 1)
```

`compliance()` computes something different: the proportion of scheduled prompts
that received a response *inside their response window*.

```{r}
cp <- compliance(
  ema,
  start_date = "2026-02-18",
  n_days = 70,
  times = c("10:00", "16:00", "22:00"),
  window = 3
)
cp
```

The two numbers differ by about three points on the same data, for two reasons.
A prompt answered three times counts once here, not three times. And a record
arriving outside every window is data, but it is not evidence that a prompt was
answered when it was asked.

Neither number is wrong. They answer different questions, and a paper reporting
"compliance" without saying which one leaves the reader unable to tell. The
prompt-level detail is available when the headline number needs unpacking:

```{r}
tapply(cp$prompts$answered, cp$prompts$time, mean)
```

Response rates were not equal across the day. Whether the pattern is large enough
to matter is a judgement about the study, but it is the kind of thing worth seeing
before treating missingness as random.

## 5. Analysis

From here the data are ordinary. `nof1kit` deliberately stops at the boundary of
analysis: intensive longitudinal data are well served by existing tools, and the
gap this package fills is upstream of them.

```{r, eval = FALSE}
library(lme4)
lmer(mood ~ melatonin + (1 | study_day), data = ema)
```

## Summary

```{r, eval = FALSE}
sched <- design_schedule(n_days = 70, max_run = 2, seed = 20260218)
write_schedule(sched, "schedule.json")     # onto the phone

ema <- read_ema("export.csv", start_date = "2026-02-18")
validate_ema(ema, ranges = list(mood = c(0, 100)), n_days = 70)
compliance(ema, start_date = "2026-02-18", n_days = 70,
           times = c("09:00", "15:00", "21:00"))
```

Four functions cover design, handoff, quality control, and compliance. What
happens after that is what the rest of the R ecosystem is for.
